VF1 Vignette

### 2024-03-12 vignette.Rmd
Re-structured text.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
```

@@

Visual field testing is a psychophysical test that measures a patient's visual ability at different locations of the visual field. This ability is usually highest in the center of the visual field (i.e. the point that the observer looks at) and decreases toward the periphery (hill of vision).

Theoretically, different aspects of vision can be tested at different locations (ability to discriminate dots that are very close to each other, ability to discriminate similar colors, contrast sensitivity), but usually white-on-white perimetry is performed where a white spot is presented on a white background.

The most common type of visual field testing is *perimetry*, which is performed with in a dome-shaped cupule, so that the distance between the stimulus and the eye equal for all locations within the visual field. In contrast, *campimetry* is performed with a flat screen.

Psychophysical tests measure a subjective perception of a physical stimulus by systematically varying the stimulus and asking the observer to communicate his perception with a limited number of allowed responses. In visual field testing, the observer usually presses a button when a light spot is perceived. Usually, psychophysical testing results in a *threshold*, which denotes either the absolute stimulus strength that can just barely be perceived or the smallest difference that can be perceived as different. Real world scenarios where perception is close to threshold include going through a dark forest on a cloudy night or seeing the car in front through a dense fog. Therefore, visual field testing is often considered as tedious by patients and judges as a less reliable examination. The latter is not correct.

The visual fields package was created to analyze data from *static perimetry*, where a threshold is measured for each of a fixed number of locations in a grid pattern. In contrast, in *kinetic perimetry*, a light spot moves from the periphery toward the center along a number of meridians and the location there the spot is first perceivable for the subject is noted.

# Starting with an example

The visualFields package can be installed from CRAN.

@@

library(visualFields)
# install.packages("visualFields", repos = "https://cloud.r-project.org")

The visualField package contains some example data.

@@

This dataset contains the results of 42 visual field tests of one patient (id: sample1). The patient has a chronic-neurodegenerative disease of the optic nerve called primary open-angle glaucoma (type: pwg).

@@

# Visual field tests

## Grid

The spatial position of the points \$l_1 \$to $ln$ in the visual field is found in the locmap and depends on the pattern used by the perimeter. When analysing data, you need to specify the locmap.

plot(x = default_locmap$coord$x, 
     y = default_locmap$coord$y, 
     type = "point", 
     asp = 1,
     xlim = c(-24, 24),
     ylim = c(-24, 24))
points(x = default_locmap$coord[default_locmap$bs, "x"],
       y = default_locmap$coord[default_locmap$bs, "y"],
       col = "red")

@@

# Differential light sensitivity and other mathematical derivatives of threshold

@@

>>>>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
library(visualFields)

example1 <- vffilter(vfpwgSunyiu24d2, 
                     id == "sample1" & 
                       eye == "OD" & 
                       date == min(vfpwgSunyiu24d2$date))

# example1_dataframe <- data.frame(x = getlocmap()$coord$x, y = getlocmap()$coord$y, td = as.vector(unlist(example1[1, 11:64])))

@@

Visual field testing measures a patient's visual ability at different locations of the visual field. This ability is usually highest in the center of the visual field (i.e. the point that the observer looks at) and decreases toward the periphery (*hill of vision*). The most common type of visual field testing is *perimetry*, which is performed with in a dome-shaped cupule, so that the distance between the stimulus and the eye equal for all locations within the visual field. In contrast, *campimetry* is performed with a flat screen.

Usually, perimetry tests the ability to perceive differences in luminance with an achromatic (not colored) stimulus (white-on-white perimetry). However, theoretically, other aspects of vision can be tested at different locations (ability to discriminate dots that are very close to each other, ability to discriminate similar colors, contrast sensitivity). The visual fields package was created to analyze data from *static perimetry*, where a threshold is measured for each of a fixed number of locations in a grid pattern. In contrast, in *kinetic perimetry*, a light spot moves from the periphery toward the center along a number of meridians and the location there the spot is first perceivable for the subject is noted.

# Performing visual field tests

## Grid pattern

The spatial position of the points $l_1$ to $l_n$ in the visual field is found in the locmap and depends on the pattern used by the perimeter. When analysing data, you need to specify the locmap.

```{r echo=FALSE}

default_locmap <- getlocmap()
plot(x = default_locmap$coord$x, 
     y = default_locmap$coord$y, 
     type = "point", 
     asp = 1,
     xlim = c(-24, 24),
     ylim = c(-24, 24))
points(x = default_locmap$coord[default_locmap$bs, "x"],
       y = default_locmap$coord[default_locmap$bs, "y"],
       col = "red")

# Clinical applications of visual field tests

In clinical practice, visual field testing is used identify decreases in visual function caused by disease. Such *visual field defects* or *scotomas* can be caused by diseases that affect tissue anywhere along the optic pathway. This ranges from the optic media (cornea / lens / vitreous) which transmit light, over the retina which detects lights and preprocesses visual information, the optic nerve which transmits information, up to the brain.

Often, clinicians can draw reliable conclusions about the localisation of the disease by the pattern of the visual field defects. For example, an infarction of the right visual cortex leads complete blindness in the left hemispheres of both eyes with a very distinct border along the vertical meridian. Therefore, the patient does not perceive anything to the right of the point they look at. On the other side of vertical meridian the visual field is normal. In contrast, patients with glaucoma start with areas of slightly perceived light sensitivity often in the upper or lower hemifield with a bow-shaped pattern cause by the localisation of damage in the optic nerve head.

Visual field tests can be used to diagnose disease, to appraise the functional consequences of disease (for example: can the patient still drive a vehicle?), and to monitor progression of disease.

The most advanced statistical methods have been created for monitoring glaucoma progression. The open-angle glaucomas are a group of slowly progressing, chronic neurodegenerative diseases of the optic nerve, often caused by increased eye pressure. Patients need consequent monitoring over many years and slight changes in the visual field must be detected reliably. Therefore, many of the functions of the visualField package focus on glaucoma research.

# Getting started with the package

The visualField package contains some example data. This dataset contains the results of 42 visual field tests of one patient (id: sample1). The patient has a chronic-neurodegenerative disease of the optic nerve called primary open-angle glaucoma (type: pwg).

@@

# install.packages("visualFields")
library(visualFields)

@@

@@

# Psychophysical testing

Psychophysical tests systematically measure the relationship between physical stimulus strength and the subjective perception. This is performed by varying the stimulus and asking the observer to communicate his perception with a limited number of allowed responses. In visual field testing, the observer usually presses a button when a light spot is perceived. The relationship between stimulus and perception is analyzed using mathematical model. Commonly, these models result in an estimate of a *threshold*, which denotes either the absolute stimulus strength that can just barely be perceived or the smallest difference that can be perceived as different. Real world scenarios where perception is close to threshold include going through a dark forest on a cloudy night or seeing the car in front through a dense fog. Stimuli are not clearly seen when close to threshold. For this reason, patients sometimes feel that lengthy visual field testing is tedious and that it is as an unreliable examination. The latter is not clearly correct when fluctuations are taken into account and data are analyzed with appropriate statistical methods.

@@

# Mathematical derivatives of threshold

## Sensitivity and other mathematical derivatives of threshold

@@

# Advanced topics

## Locini and Number of locations / vfisvalid

## Locmaps
